---
title: "Tutorial topic 8, Simple and Multiple linear regression"
date: "Document generated on `r Sys.Date()`"
output: 
    pdf_document:
      fig_width: 7
      fig_height: 6
      fig_caption: true
---
```{r setup, warning=F, message=F, echo=F}
# root dir
knitr::opts_knit$set(root.dir = 
                       "z:/willem/teaching/envx2001/otherdata")
library(pander)
library(knitr)
```

# Objectives  

*  Review calculating (i) ANOVA table of a simple linear regression model (ii) the correlation coefficient;
*  Learn to interpret correlation matrix to identify useful predictor variables;
*  Learn to calculate regression parameters for a multiple linear regression model.

## EXERCISE 1 – REVIEW: FITTING REGRESSION PARAMETERS & ASSESSING SIGNIFICANCE OF A SIMPLE LINEAR REGRESSION MODEL

The number of horses on Canadian farms appeared to decrease after the war:
```{r data, echo=F}
pander(t(data.frame(Year = c(1944, 1945, 1946, 1947,	1948),
"horses_mult_100,000" = c(28,	26,	22,	20,	19))))
```

a)	Calculate the parameters of the line of best fit (simple linear regression). Use slide 6 - 8 from the lectures in week 8 to assist you



```{r emptyTable1, echo=F}
pander(data.frame("data" = c(rep("",5),"mean"),Year = c(1944, 1945, 1946, 1947,	1948,""), 	
                  Horses =	c(28,	26,	22,	20,	19, ""), "Year-mean" = rep("",6),	"Horses-mean" = rep("",6),	"Product" = rep("",6)))
```			


LINE OF BEST FIT: number of horses  =  		year

b)	Complete the following hypotheses and ANOVA:

$H_0$:	
$H_1$:


```{r ANOVAtable, echo=F}

pander(data.frame(Source_of_variation = c("Regression", "Residual", "Total"), 	df = rep("",3),	SS = c(57.6,"",60.0),	MS = rep("",3),
                  F = rep("",3),	p_value = rep("",3)))
```


c)	Is the model significant? Explain why

##EXERCISE 2 – CORRELATION MATRIX

A biological survey was performed to examine the nature of the variables thought to influence the breeding of British dippers (thrush-sized birds, living mainly in the upper reaches of rivers; they feed on benthic invertebrates by probing the river beds with their beaks). Twenty two sites were included in the survey.  The predictor variables that were measured are:  

•	site altitude  
•	water hardness  
•	river-bed slope  
•	the numbers of caddis fly larvae  
•	the numbers of stonefly larvae  
•	the numbers of mayfly larvae  
•	the numbers of all other invertebrates collected  
•	the breeding density which is the number of breeding pairs of dippers per 10 km of river (this is the response variable)   

In the analyses, we transform the four invertebrate variables using a log10(number+1) transformation.

A correlation matrix presents the Pearson correlation coefficients.

```{r corMatrix, echo=F}
pander(t(data.frame("variable" =	c("Altitude","Hardness",
                                   "RiverSlope",	"LogCadd",
                               "LogStone",	"LogMay",	"LogOther"),
      Hardness =	c(-0.082,rep("",6)),
      RiverSlope =	c(0.490,	0.242, rep("",5)),
      LogCadd = c(0.181,	0.334,	0.431, rep("",4)),
      LogStone = c(0.477,	0.030,	0.575,	0.443, rep("",3)),
      LogMay =	c(-0.128,	0.414,	-0.096,	0.438,	-0.167,rep("",2)),
     LogOther = c(-0.339,	0.554,	-0.020,	0.352,	-0.300,	0.388, ""),
      Br.Dens = c(0.406,	0.350,	0.710,	0.613,	0.763,	0.156,
                  -0.128))))
```

a)	Why would we have transformed the counts of invertebrates?  
b)	Why would add 1 to the count before performing the log transformation?  
c)	What is the best single predictor of breeding density? What is the second best single predictor? What is the worst? Had the correlation between Br Dens and LogStone been 
-0.763, would any of these conclusions change?  
d)	If you had to choose two predictors to use in a MLR model, which would you choose?  

 
## EXERCISE 3 – CALCULATION OF REGRESSION PARAMETERS
Please look at the lecture slides 29 and 30 from Topic 8 to assist you

The variance-covariance matrix for these variables is:
```{r vcovMatrix, echo=F}
pander(t(data.frame("variable" =	c("Altitude","Hardness",
                                   "RiverSlope",	"LogCadd",
                               "LogStone",	"LogMay",	"LogOther", "Br.Dens"),
      Altitude =	c(3143.9,	-196.8,	172.1,	9.4,	26.7,	-13.4,
                   -34.7,	33.9),
       Hardness =	c(-196.8,	1845.1,	65.1,	13.4,	1.3,	33.2,	43.4,
                    22.4),
      RiverSlope =	c(172.1,	65.1,	39.2,	2.5,	3.6,	-1.1,	-0.2,
                     6.6),
      LogCadd = c(9.4,	13.4,	2.5,	0.9,	0.4,	0.8,	0.6,	0.8),
      LogStone = c(26.7,	1.3,	3.6,	0.4,	1.0,	-0.3,	-0.5,	1.1),
      LogMay =	c(-13.4,	33.2,	-1.1,	0.8,	-0.3,	3.5,	1.3,	0.4),
     LogOther = c(-34.7,	43.4,	-0.2,	0.6,	-0.5,	1.3,	3.3,	-0.3),
      Br.Dens = c(33.9,	22.4,	6.6,	0.8,	1.1,	0.4,	-0.3,	2.2))))
```


The mean values are 

(i) breeding density = 4.7 
(ii) RiverSlope = 9.9 
(iii) log(stonefly numbers +1) = 5.0

a)	Write down the equations you would solve to obtain the parameters for the line of best fit for predicting breeding density using the two predictors RiverSlope and log(stonefly numbers +1).  

Line of best fit is:  
\begin{center}
$breeding_density = b_0 + b_1 RiverSlope + b_2 log(stonefly numbers +1)$.
\end{center}

b)	**If time permits** solve the simultaneous equations to estimate b_1, b_2 and then b_0.

